Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model

### 2021-10-19 manuscripts/owp_fim3_2020/sections/materials_and_methods.tex
M&M and eval_plots updates
Updates * first draft complete * version sent out to JJ

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
FIM 3 is a fully operational pipeline of software tools to help acquire datasets, cache hydrofabrics, produce FIMs, and evaluate results.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software Dependencies and Architecture}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM exclusively utilizes free and open source software dependencies including Python 3, GDAL, TauDEM, Geographic Resource Analysis Support System (GRASS), GNU Parallel, and MPICH \cite{python382,gdal2020,tarboton2005terrain,grass2020,tange2015gnu,amer2021mpich}.
Within the Python 3 ecosystem, many common packages are employed including but not limited to RichDEM, GeoPandas, Rasterio, Rasterstats, and Numba \cite{barnes2018richdem,jordahl2014geopandas,lam2015numba}. 
To simplify setup and enhance portability across host operating systems OWP FIM packages all dependencies up in Docker image (\url{https://docs.docker.com/engine/install/}). 
A user only need to install Docker on their host machine and build the image from the provided recipe.

The FIM 3 pipeline is discretized into key areas that a user can interact with to reproduce the results of this study. Preprocessing acquires and prepares datasets for production of the FIM 3 hydrofabric. 
Producing the FIM hydrofabric produces the datasets required to make an inundation map including the relative elevation model (REM) or HAND grid, the catchments in vector and raster form, and the hydro-table (contains synthetic rating curves and cross-walk information).
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\label{ssec:datasets}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Most data sources used within FIM 3 are publicly available from a variety of government sources including the USGS, NWC, Federal Emergency Management Agency (FEMA), and US Army Core of Engineers (USACE) to enhance reproducibility and collaboration among government, academia, and industry.
The National Hydrography Dataset Plus High Resolution (NHDPlusHR) Beta Version is the latest hydrography dataset used for land surface hydrologic modeling in the US \cite{moore2019user}. 
It is used in conjunction with the hydrofabric of the NWM V2.1 to help define flowlines for FIM 3 while the NWM hydrofabric is also used to define reservoirs for exclusion and catchments to cross-walk to for forecasting purposes.
For enforcing levee data into the NED DEMs, the USACE National Levee Database (NLD) is used via burning feature elevations \cite{engineers2016national}.
Additionally, FEMA Base Level Engineering (BLE) from Region 6 (parts of Texas, Oklahoma, Arkansas, Louisiana and New Mexico) 1\% (100 year) and 0.2\% (500 year) datasets are used as validation in this study \cite{fema2021base,fema2021estimated}. 
These BLE datasets are provided at the watershed scale (HUC8) utilizing best available simulations and DEMs.
The full input datasets presented by source are list in Table \ref{tab:data}.
\note[Fernando Aristizabal]{Check Brian Avant, Trevor Grout, Bradford Bates, Ryan Spies}
%

\begin{table}
\caption{Data sources, names, and descriptions.}
\label{tab:data}
\centering
\begin{tabular}{|p{2cm}|p{4cm}|p{8cm}|}
%\begin{tabular}{l c c}
\hline
Source & Name & Description \\
\hline
USGS & NHDPlusHR BurnLineEvents & Stream lines used by NHDPlus HR for hydro-enforcement \\
\hline
USGS & NHDPlusHR Value-Added Attributes & Database of additional attributes associated with the BurnLineEvents that enhance navigation, analysis, and display \\
\hline
USGS & NHDPlusHR DEM & DEM used for NHDPlus HR at 1/3 arc-second (10m) spatial resolution and vertical units in centimeters \\
\hline
NOAA-OWP & NWM Streams & Stream network center lines used by NWM for routing and forecasting \\
\hline
NOAA-OWP & NWM Catchments & Surface drainage area corresponding to each reach in the NWM \\
\hline
NOAA-OWP & NWM Waterbodies & Waterbodies considered by the NWM as resevoirs or lakes \\
\hline
USACE & NLD & Levee database of locations and elevations  \\
\hline
WHERE? & Land-Sea Border & Border of land and sea and land and Great Lakes \\
\hline
FEMA & Cross-Sections & HEC-RAS cross-sections used for modeling in BLE datasets. Include discharges for 1\% and 0.2\% recurrence interval events \\
\hline
FEMA & Flood Inundation Extents & Extents produced by FEMA BLE for 1\% and 0.2\% recurrence interval events \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

Areas with all the required data (from the NWM and the USGS) are labeled as the FIM domain. 
\note[Fernando Aristizabal]{For Nick Chadwich} This includes XXXX HUC4's, XXX HUC'6s, and XX HUC8's.
An enhancement of OWP FIM v2 `Cahaba' over previous FIM versions is the support for Hawaii and Puerto Rico which the NWM v2.1 will cover.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Hydro-conditioning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The DEM from the NED is subject to a series of hydro-conditioning procedures to enhance it's suitability for riverine flood inundation mapping. 
These techniques are specific for making OWP FIM and differ from the conditioning methods used by the NHDPlusHR Beta \cite{moore2019user}.
Hydro-conditioning is implemented to obtain many objectives including enforcing the location of hydrologically relevant features such as flowlines, lakes, or drainage divides whether natural or anthropogenic. 
It can also be used to simulate more accurate bathymetry which is not accounted for in the NED 10m DEM \cite{gesch2002national}.

Specifically within the context of FIM 3, the hydro-conditioning operations that take place in sequential order are presented. 
Prior to any hydro-conditioning, all input datasets must be subset from their original spatial domain scales into the processing unit designated at run time which can be either HUC 4, 6 or 8. 
The subsetting is done by spatial query for the cases of the levees, DEM, and NWM hydrofabric while the NHDPlusHR BurnLineEvents are subset via attribute query for the given reachcode's membership in the processing unit.
Hydro-conditioning raster operations take place on buffered boundary definitions to avoid edge contamination and effects \cite{lindsay2013measuring}. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Stream Network Enforcement} 
\label{ssec:stream_network_enforcment}
%
\note[Fernando Aristizabal]{For Brian Avant}
The location of the stream network are enforced to ensure general agreement with established stream networks.
The NHDPlus HR Beta Burnline Event layer is used to enforce stream locations in the NHDPlusHR workflow so it is also used here for hydro-enforcement \cite{moore2019user}. 
However, to better match the drainage density of the NWM V2.1 stream network which is based on the NHDPlus Medium Resolution, the Burnline Events are pruned utilizing a nearest neighbor search around the NWM flowlines. 
Pruning the NHD network is completed at for every NWM V2.1 headwater point, the nearest NHDPlusHR Burline Event feature is selected and downstream features are also selected.
The resulting pruned NHD stream network is what gets hydro-enforced in subsequent operations.
This procedure is best illustrated in Figure \ref{fig:stream_density_pruning} which shows that the pruned NHD network corresponds to the full density NHD network at NWM V2.1 headwater locations only. 
Additionally, the NHDPlusHR pruned headwaters are later used for seeding a new FIM drainage network that best agrees withe DEM after all hydro-conditioning takes place.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/headwaters.jpg}
\caption{Pruning of NHDPlus HR Beta Burnline Events (dotted black) to NWM V2.1 stream density (blue) using nearest neighbor selection. Resulting stream network (red) matches the drainage density of NWM V2.1 while corresponding spatially with the NHDPlusHR Burnline Events.}
\label{fig:stream_density_pruning}
\end{figure}
%
This results in a stream network that has the same density as the NWM V2.1 flowline network but utilizes the locations of the NHDPlusHR Beta BurnLineEvents.

@@

@@

The depression technique employed here does leave the existence of flat regions where pits existed aprior thus later requiring the need for resolving these flats.

@@

In order to prevent situations where the burned thalweg and the thalweg endemic to DEM run parallel to one another, the normalized excavation algorithm \cite{saunders1999preparation} is used to seek a zonal (nearest neighbor) elevation minimum for each thalweg pixel. 

@@

The next step involves conditioning these local minimums along the thalweg to enforce monotonically decreasing thalweg elevations for FIM.
\citeA{garousi2019terrain} proposed an algorithm that traverses stream thalweg pixels in a depth first manner starting with adding all the headwater pixels to a queue. 
The connectivity of the thalweg pixels is defined by the D-8 flow directions further discussed in Section \ref{ssec:flow_direction_and_flat_resolution}.
At every thalweg pixel, the minimum elevation among itself and its upstream contributing thalweg pixels is taken as shown in equation \ref{eq:thalweg_breach},
%
\begin{equation}
\label{eq:thalweg_breach}
\textbf{D}[x] = \min_{y\ drains\ to\ x} {(\ \textbf{D}[x]\ ,\ \textbf{D}[y]\ )}
\end{equation}
%
, in which \textbf{D} represents the array of thalweg adjusted elevations indexed by x and y where by y is upstream of x. 
When a pixel's upstream neighbors are all evaluated, the downstream pixel is added into the queue thus the depth first traversal of the drainage network.
This procedure enforces the location of streams and ensures that thalweg elevations are hydrologically correct which yielded a 7\% improvement in critical success index (CSI) per an evaluation for an event in 2017 on the Malad river \cite{garousi2019terrain}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deriving FIM Hydrofabric}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The FIM Hydrofabric is defined here as the collection of geospatial datasets that are used for converting NWM discharges into inundation extents.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Flow Directions and Flats Resolution}
\label{ssec:flow_direction_and_flat_resolution}
%
To facilitate the generation of a connected stream network and its associated catchments from the conditioned DEM, the depression-filled DEM is used to derive connectivity in the form of D-8 flow directions.
D-8 seeks to allocate a drainage direction for every pixel based on the adjacent eight pixel neighborhood with the steepest slope \cite{o1984extraction}.
The horizontal component of slope is defined as one for the 4 neighboring pixels in the main cardinal directions while the intercardinal pixels are designated a horizontal component of $\sqrt{2}$. 
The flow direction is encoded as integers 1 through 8 corresponding with the cardinal direction East as 1 and continuing counter-clockwise to the Southwest direction as 8. 
Flow directions are derived for non-depression filled regions trivially with the above procedure but to define connectivity for every grid cell the remaining flats corresponding to depression filled cells must be resolved.

@@

The FIM stream network is of the same density as the NWM V2.1 network and fully converges at all junctions leaving no divergences in the network.

@@

Based on previous research, splitting each remaining stream link into equi-distant reaches not to exceed a parameterized value of 1.5km helps improve synthetic rating curve and mapping skill \cite{garousi2019terrain,godbout2019error,zheng2018geoflood}.
Small reaches can lead to unrealistic variances in channel geometries while oversized reaches can lead to grouping too much geometry variance into one parametrized unit.

@@

@@

Every non-thalweg elevation is substracted from the thalweg elevation within the same pixel-level catchment.

@@

The reach average Manning's formula is derived to be

\begin{equation}
\label{eq:reach_averaged_mannings_equation}
\textbf{Q}[y] = \frac{1}{n} \frac{V[y]^{5/3}S^{1/2}}{L B[y]^{2/3}} 
\end{equation}

where Q is discharge, y indicates the stage, n is the Manning's n roughness coefficient, V is volume at stage y, S is channel slope, L is along flow length, and B is wetted bed area at stage y.
All units are international given the 1 numerator above n.
The reach averaged method has been compared to rating curves from Hydrologic Engineering Center\'s River Analysis System (HEC-RAS) and USGS gages yielding comparable results for estimating the river bottom elevation profile, channel width at given stages, and stage-discharge relationships \cite{zheng2018river}.
The reach averaged geometry parameters including number of cells, bed area, and volume are sampled from the thalweg conditioned AGREE DEM using TauDEM's catchhydrogeo utility.
Using the split reaches described in Section \ref{sssec:deriving_fim_stream_network}, the channel slope is sampled from the thalweg conditioned DEM at the end points of the reaches while the same reaches are used to the channel length.
Calibration of the Manning's n roughness coefficient is conducted by optimizing inundation extent performance to a benchmark dataset best described in Section \ref{ssec:evaluation_and_calibration}. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Unary Stream Order}
\label{ssec:unary_stream_order}
%
FIM skill has been shown to be sensitive to the drainage density of the stream network employed as the datum for HAND \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
In our evaluations, we note negative effects at the confluence of lower flow tributaries with higher flow rivers partly due to the independent nature of the catchments within HAND methods.
Figure \ref{fig:confluence_problem_statement} illustrates this exact situation where two tributatries converge with a higher order stream segment. 
A `Cahaba' FIM is generated and compared with a FEMA 100 year extent (please see sections \ref{ssec:evaluation_and_calibration} for more details) showing significant under-prediction in inundation extent
The higher discharge along the mainstem of 1,900 cubic meters per second (CMS) does not translate to the lower flow rates along the tributaries of 84 and 195 CMS. 
This is due to a lack of representation of backwater conditions in the hydraulic routing techniques used.
As a parallel problem, there is excess water accumulated along the mainstem that cannot extend in either a fluvial or pluvial manner beyond the boundaries of the mainstem catchments.

\begin{figure}[h!]
\centering
%\includegraphics[scale=1.0]{figures/headwaters.jpg}
\caption{}
\label{fig:confluence_problem_statement}
\end{figure}
%
The limitation imposed by using stream networks of Horton-Strahler stream order \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} greater than one for HAND can be described mathematically with Graph Theory.
Assume a drainage area denoted by special case of a fully connected, directed acyclic graph known as a tree, G, with all nodes (or reaches) directed from the leaf nodes (or headwater reaches) toward a singular root node $r_0$ known as the outlet in the application's field.
Assume a singluar subgraph of G where by a new root node,$r_n$, is introduced that represents another stream location.
n is used here to index possible root nodes which are a subset of G.
Ensuring $r_n$ is itself an inner node (has upstream drainage) disconnects G into $G_0$ and $G_n$ thus removing influence of conditions at $r_0$ upon $G_n$.
HAND seeks to divide drainage areas into reach scale catchments that are independent of one another.
We intend to demonstrate how introduction of stream orders greater than one can generate errors at confluences.
We present two successive methods implemented that reduce drainage densities by reducing Horton-Strahler stream orders of the networks employed and test our presented hypothesis that unary stream order networks enhance FIM performance skill with HAND.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{NWS Replace and Route Mainstems}
\label{ssec:unary_stream_order}
%

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Generalized Mainstems}
\label{ssec:unary_stream_order}
%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
FIM 3 is a fully operational pipeline of software tools to help acquire datasets, cache hydrofabrics, produce FIMs, and evaluate results.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Software Dependencies and Architecture}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
OWP FIM exclusively utilizes free and open source software dependencies including Python 3, GDAL, TauDEM, Geographic Resource Analysis Support System (GRASS), GNU Parallel, and MPICH \cite{python382,gdal2020,tarboton2005terrain,grass2020,tange2015gnu,amer2021mpich}.
Within the Python 3 ecosystem, many common packages are employed including but not limited to RichDEM, GeoPandas, Rasterio, Rasterstats, and Numba \cite{barnes2018richdem,jordahl2014geopandas,lam2015numba}. 
To simplify setup and enhance portability across host operating systems OWP FIM packages all dependencies up in Docker image (\url{https://docs.docker.com/engine/install/}). 
A user only need to install Docker on their host machine and build the image from the provided recipe. 
Source code is made available for this project on GitHub (https://github.com/NOAA-OWP/cahaba).

The FIM 3 pipeline is discretized into key areas that a user can interact with to reproduce the results of this study. Preprocessing acquires and prepares datasets for production of the FIM 3 hydrofabric. 
Producing the FIM hydrofabric produces the datasets required to make an inundation map including the relative elevation model (REM) or HAND grid, the catchments in vector and raster form, and the hydro-table (contains synthetic rating curves and cross-walk information).
A user should visit the Readme.md page on GitHub for more information on how to acquire the datasets and reproduce the pipeline.
\subsection{Datasets}
\label{ssec:datasets}
Most data sources used within FIM 3 are publicly available from a variety of government sources including the USGS, NWC, Federal Emergency Management Agency (FEMA), and US Army Core of Engineers (USACE) to enhance reproducibility and collaboration among government, academia, and industry.
Instructions for accessing data are provided on the projects GitHub page via an Amazon Web Services (AWS) S3 bucket.
The National Hydrography Dataset Plus High Resolution (NHDPlusHR) Beta Version is the latest hydrography dataset used for land surface hydrologic modeling in the US \cite{moore2019user}. 
It is used in conjunction with the hydrofabric of the NWM V2.1 to help define flowlines for FIM 3 while the NWM hydrofabric is also used to define reservoirs for exclusion and catchments to cross-walk to for forecasting purposes.
For enforcing levee data into the NED DEMs, the USACE National Levee Database (NLD) is used via burning feature elevations \cite{engineers2016national}.
Additionally, FEMA Base Level Engineering (BLE) from Region 6 (parts of 9 states including Colorado, New Mexico, Texas, Oklahoma, Kansas, Arkansas, Louisiana, Missouri and Mississippi) 1\% (100 year) and 0.2\% (500 year) recurrence flow datasets are used as validation in this study \cite{fema2021base,fema2021estimated}. 
These BLE datasets are provided at the watershed scale (HUC8) utilizing best available simulations and DEMs.
The full input datasets presented by source are list in Table \ref{tab:data}.
\note[Fernando Aristizabal]{Check Brian Avant, Trevor Grout, Bradford Bates, Ryan Spies}
\begin{table}
\caption{Data sources, names, and descriptions.}
\label{tab:data}
\centering
\begin{tabular}{|p{2cm}|p{4cm}|p{8cm}|}
%\begin{tabular}{l c c}
\hline
Source & Name & Description \\
\hline
USGS & NHDPlusHR BurnLineEvents & Stream lines used by NHDPlus HR for hydro-enforcement \\
\hline
USGS & NHDPlusHR Value-Added Attributes & Database of additional attributes associated with the BurnLineEvents that enhance navigation, analysis, and display \\
\hline
USGS & NHDPlusHR DEM & DEM used for NHDPlus HR at 1/3 arc-second (10m) spatial resolution and vertical units in centimeters \\
\hline
NOAA-OWP & NWM Streams & Stream network center lines used by NWM for routing and forecasting \\
\hline
NOAA-OWP & NWM Catchments & Surface drainage area corresponding to each reach in the NWM \\
\hline
NOAA-OWP & NWM Waterbodies & Waterbodies considered by the NWM as reservoirs or lakes \\
\hline
USACE & NLD & Levee database of locations and elevations  \\
\hline
WHERE? & Land-Sea Border & Border of land and sea and land and Great Lakes \\
\hline
FEMA & Cross-Sections & HEC-RAS cross-sections used for modeling in BLE datasets. Include discharges for 1\% and 0.2\% recurrence interval events \\
\hline
FEMA & Flood Inundation Extents & Extents produced by FEMA BLE for 1\% and 0.2\% recurrence interval events \\
\hline
\end{tabular}
\end{table}
Areas with all the required data (from the NWM and the USGS) are labeled as the FIM domain which includes 2,188 HUC8s for the FR and Generalized Mainstems (GMS) networks and 1,604 HUC8s for the MS method. 
These methods will be explained more later.
An enhancement of OWP FIM `Cahaba' over previous HAND based FIM versions is the support for Hawaii and Puerto Rico which the NWM v2.1 will cover.
\subsection{Hydro-conditioning}
\label{ssec:hydro_conditioning}
The DEM from the NED is subject to a series of hydro-conditioning procedures to enhance it's suitability for riverine flood inundation mapping. 
These techniques are specific for making OWP FIM and differ from the conditioning methods used by the NHDPlusHR Beta \cite{moore2019user}.
HAND inherently requires all areas eligible for inundation to drain to the designated drainage network so DEMs must undergo significant manipulation to make this the case.
Hydro-conditioning is implemented to obtain many objectives including enforcing the location of hydrologically relevant features such as flowlines, lakes, or drainage divides whether natural or anthropogenic. 
It can also be used to simulate more accurate bathymetry which is not accounted for in the NED 10m DEM \cite{gesch2002national}.
HAND inherently requires all areas eligible for inundation to drain to the designated drainage network so DEMs must undergo significant manipulation to make this the case.

Specifically within the context of FIM 3, the hydro-conditioning operations that take place in sequential order are presented. 
Prior to any hydro-conditioning, all input datasets must be subset from their original spatial domain scales into the processing units of size HUC8. 
The subsetting is done by spatial query for the cases of the levees, DEM, and NWM hydrofabric while the NHDPlusHR BurnLineEvents are subset via attribute query for the given reach code's membership in the processing unit.
Hydro-conditioning raster operations take place on buffered boundary definitions to avoid edge contamination and effects \cite{lindsay2013measuring}. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Stream Network Enforcement} 
\label{ssec:stream_network_enforcment}
%
\note[Fernando Aristizabal]{For Brian Avant}
The location of the stream network is enforced to ensure general agreement with established stream networks.
The NHDPlus HR Beta Burnline Event layer is used to enforce stream locations in the NHDPlusHR workflow so it is also used here for hydro-enforcement \cite{moore2019user}. 
However, to better match the drainage density of the NWM V2.1 stream network which is based on the NHDPlus Medium Resolution, the Burnline Events are pruned utilizing a nearest neighbor search around the NWM flowlines. 
Pruning the NHD network is completed at for every NWM V2.1 headwater point where the nearest NHDPlusHR Burnline Event feature is selected and downstream features are also selected.
The resulting pruned NHD stream network is what gets hydro-enforced in subsequent operations.
This procedure is best illustrated in Figure \ref{fig:stream_density_pruning} which shows that the pruned NHD network corresponds to the full density NHD network at NWM V2.1 headwater locations only. 
Additionally, the NHDPlusHR pruned headwaters are later used for seeding a new FIM drainage network that best agrees with the DEM after all hydro-conditioning takes place.
%
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/headwaters.jpg}
\caption{Pruning of NHDPlus HR Beta Burnline Events (dotted black) to NWM V2.1 stream density (blue) using nearest neighbor selection. Resulting stream network (red) matches the drainage density of NWM V2.1 while corresponding spatially with the NHDPlusHR Burnline Events.}
\label{fig:stream_density_pruning}
\end{figure}
%
This results in a stream network that has the same density as the NWM V2.1 flowline network but utilizes the locations of the NHDPlusHR Beta BurnLineEvents.

@@

The elevations found in the NLD are burned into the DEM if those elevations were found to exceed those already in place.

@@

The depression technique employed here does leave the existence of flat regions where pits existed a prior thus later requiring the need for resolving these flats.

@@

In order to prevent situations where the burned thalweg and the thalweg endemic to the DEM run parallel to one another, the normalized excavation algorithm \cite{saunders1999preparation} is used to seek a zonal (nearest neighbor) elevation minimum for each thalweg pixel. 

@@

The next step involves conditioning these local minimums along the thalweg to enforce monotonically decreasing thalweg elevations for FIM.
\citeA{garousi2019terrain} proposed an algorithm that traverses stream thalweg pixels in a depth first manner starting with adding all the headwater pixels to a queue. 
The connectivity of the thalweg pixels is defined by the D-8 flow directions further discussed in Section \ref{ssec:flow_direction_and_flat_resolution}.
At every thalweg pixel, the minimum elevation among itself and its upstream contributing thalweg pixels is taken as shown in Equation \ref{eq:thalweg_breach},
%
\begin{linenomath*}
\begin{equation}
\label{eq:thalweg_breach}
\textbf{D}[x] = \min_{y\ drains\ to\ x} {(\ \textbf{D}[x]\ ,\ \textbf{D}[y]\ )}
\end{equation}
\end{linenomath*}
%
, in which \textbf{D} represents the array of thalweg adjusted elevations indexed by x and y where by y is upstream of x. 
When a pixel's upstream neighbors are all evaluated, the downstream pixel is added into the queue thus the depth first traversal of the drainage network.
This procedure enforces the location of streams and ensures that thalweg elevations are hydrologically correct which yielded a 7\% improvement in critical success index (CSI) per an evaluation for an event in 2017 on the Malad river \cite{garousi2019terrain}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deriving FIM Hydrofabric}
\label{ssec:deriving_fim_hydrofabric}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The FIM Hydrofabric is defined here as the collection of geospatial datasets that are used for converting NWM discharges into inundation extents.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Flow Directions and Flats Resolution}
\label{ssec:flow_direction_and_flat_resolution}
%
To facilitate the generation of a connected stream network and its associated catchments from the conditioned DEM, the depression-filled DEM is used to derive connectivity in the form of D-8 flow directions.
D-8 seeks to allocate a drainage direction for every pixel based on the adjacent eight pixel neighborhood with the steepest slope \cite{o1984extraction}.
The horizontal component of slope is defined as one for the 4 neighboring pixels in the main cardinal directions while the intercardinal pixels are designated a horizontal component of $\sqrt{2}$. 
The flow direction is encoded as integers 1 through 8 corresponding with the cardinal direction East as 1 and continuing counter-clockwise to the Southwest direction as 8. 
Flow directions are derived for non-depression filled regions trivially with the above procedure but to define connectivity for every grid cell the remaining flats corresponding to depression filled cells must be resolved.

@@

The FIM stream network is of similar drainage density as the NWM V2.1 network but fully converges at all junctions leaving no divergences in the network.

@@

Based on previous research, splitting each remaining stream link into equidistant reaches not to exceed a parameterized value of 1.5km helps improve synthetic rating curve and mapping skill \cite{garousi2019terrain,godbout2019error,zheng2018geoflood}.
Small reaches can lead to unrealistic variances in channel geometries while oversized reaches can lead to grouping too much geometry variance into one parameterized unit.

@@

\label{sssec:catchments}

@@

Every non-thalweg elevation is subtracted from the thalweg elevation within the same pixel-level catchment.

@@

The reach averaged Manning's formula is derived to be 
\begin{linenomath*}
\begin{equation}
\label{eq:reach_averaged_mannings_equation}
\textbf{Q}[y] = \frac{1}{n} \frac{V[y]^{5/3}S^{1/2}}{L B[y]^{2/3}} 
\end{equation}
\end{linenomath*}
where Q is discharge, y indicates the stage, n is the Manning's N roughness coefficient, V is volume at stage y, S is channel slope, L is along flow length, and B is wetted bed area at stage y.
All units are international given the 1 numerator above n.
The reach averaged method has been compared to rating curves from Hydrologic Engineering Center\'s River Analysis System (HEC-RAS) and USGS gages yielding comparable results for estimating the river bottom elevation profile, channel width at given stages, and stage-discharge relationships \cite{zheng2018river}.
The reach averaged geometry parameters including number of cells, bed area, and volume are sampled from the thalweg conditioned AGREE DEM using TauDEM's catchhydrogeo utility.
Using the split reaches described in Section \ref{sssec:deriving_fim_stream_network}, the channel slope is sampled from the thalweg conditioned DEM at the end points of the reaches while the same reaches are used to the channel length.

Setting of the Manning's N roughness coefficient has precedent in previous CFIM studies \cite{maidment2017conceptual,liu2016cybergis,liu2020height,djokic2019arc,garousi2019terrain,zheng2018geoflood} with two noted values of 0.05 and 0.06 for NFIE and \citeA{djokic2019arc} respectively. 
These values are applied universally to the entire forecasting domain across space, time, and discharge profiles.
We note significant opportunity to enhance CFIM skill by better parameterizing Manning's N according to available data including but limited to land cover, land use, stream order, stream geometry, drainage area, reach length, and discharge percentiles \cite{garousi2019terrain,johnson2019integrated}.
For now and for the purpose of this study, we examine the developed ecosystem of tools with Manning's N set to both 0.06 and 0.12 which we hope will shed some light on the sensitivity of this parameter to HAND based FIMs.
After all the parameters to the Manning's equation have been determined with either through hydrofabric sampling or user parameterization, we select stage values of 0 to 25 meters in third of a meter increments to calculate the discharge values for each stage value. 
\subsubsection{Cross-walking NWM Networks}
\label{sssec:cross_walking_networks}
The stream network derived in Section \ref{sssec:deriving_fim_stream_network} must be associated with a NWM reach identifier so that a discharge can be converted to stage and later inundation extent.
For the methods already discussed, we overlap the reach catchments derived in Section \ref{sssec:catchments} with the NWM catchments matching the ID of the NWM catchment that most overlaps the derived catchment for HAND.
For two subsequent methods discussed in Sections \ref{sssec:replace_and_route_mainstems} and \ref{sssec:generalized_mainstems}, we find the mid-point of the derived stream reach line described in Section \ref{sssec:deriving_fim_stream_network} and find the NWM catchment that contains the mid-point.
Additionally, only relevant catchments from the NWM for the given level path are selected for cross-walking for methods in Sections \ref{sssec:replace_and_route_mainstems} and \ref{sssec:generalized_mainstems}.
While these conflation methods are approximate, they work for many instances just fine but do lead to areas with substantial error. 
More discussion on this will follow in Section \ref{sec:dicussion}.
\subsection{Stream Order Reduction}
\label{ssec:stream_order_reduction}
FIM skill has been shown to be sensitive to the drainage density of the stream network employed as the datum for HAND \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
In our evaluations, we note negative effects at the confluence of lower flow tributaries with higher flow rivers partly due to the independent nature of the catchments within HAND methods.
Figure \ref{fig:catchment_boundaries_issue} illustrates this exact situation where two tributaries converge with a higher order stream segment. 
A OWP FIM `Cahaba' FIM is generated and compared with a FEMA 100 year extent (please see sections \ref{ssec:evaluation} for more details) showing significant under-prediction in inundation extent.
The higher discharge along the mainstem of 1,900 cubic meters per second (CMS) does not translate to the lower flow rates along the tributaries of 84 and 195 CMS. 
This is due to a lack of representation of backwater conditions in the hydraulic routing techniques used.
As a parallel problem, there is excess water accumulated along the mainstem that cannot extend in either a fluvial or pluvial manner beyond the boundaries of the mainstem catchments.
\begin{figure}[h!]
\centering
\includegraphics[scale=1.0]{figures/catchment_boundaries_issue.jpg}
\caption{False negatives associated with confluence of tributaries with mainstems. Integers represent flow values from BLE 100 year for the associated areas. No backwater consideration is implemented and the independent nature of the HAND catchments prohibits pluvial inundation from taking place.}
\label{fig:catchment_boundaries_issue}
\end{figure}
We seek to resolve this problem by deriving HAND for processing units with stream networks of reduced stream order. 
We present two successive methods implemented that reduce drainage densities by reducing Horton-Strahler stream orders of the networks employed and test our presented hypothesis that unary stream order networks enhance FIM performance skill with HAND.
The resulting FIMs from the overlapping HAND processing units are mosaiced together taking an inundated area to be so.
\subsubsection{NWS Replace and Route Mainstems}
\label{sssec:replace_and_route_mainstems}
The RnR configuration of the NWM seeks to replace NWM forecasts by assimilating best available AHPS forecasts and routing them downstream.
The RnR Mainstems (MS) network is a subset of the NWM full-resolution (FR) network at and downstream of AHPS forecast points as seen in Figure \ref{fig:forecast_points}.
The RnR MS network, or simply MS, comprises about 200 thousand km of stream length which is less than 4\% of the FR total stream length of 5.5 million km.
It also spans 121,724 reaches across 1,608 HUC8s.
HAND was originally derived for this stream network to enhance mapping skill along these critical MS segments \cite{djokic2019arc}. 
The inundation derived from this stream network is mosaiced with the inundation from the FR network to form the MS FIMs. 
Within each HUC, you'll typically only find a MS stream network of stream order 1 but this can vary if more than one AHPS forecasting point is found within or upstream of the HUC in question.
\subsubsection{Generalized Mainstems}
\label{sssec:generalized_mainstems}
To further the efforts implemented by MS, we sought to derive HAND at a level-path scale which we call Generalized Mainstems (GMS).
Since the MS network only covers a small percentage of the NWM forecasting domain, we sought to expand the benefits of stream order reduction within HAND processing units to the entire FR domain.
Level-paths group flowlines by maximizing the length of each flow path and minimizing the number of level path identifiers within a given domain \cite{moore2019user,mckay2012nhdplus}. 
Starting at the outlet, a unique level-path is propagated upstream. 
At every confluence, the direction of maximum flow path length is sought to propagate the current level-path identifier.
For the remaining parent reaches of the given junction, a new level-path identifier is assigned and the process recursively continues with them.
